\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-06-09 nobuild/jss/sparseHessianFD.Rnw
jss

\documentclass[article]{jss}

@@

  \Abstract{
    Sparse Hessian matrices occur often
    in statistics, and their fast and accurate estimation can
    improve efficiency of numerical optimization and sampling
    algorithms. By exploiting the known sparsity pattern of a Hessian, methods in the
    sparseHessianFD package require many fewer function or gradient
    evaluations than would be required if the Hessian were treated as
    dense. The package implements established graph coloring and
    linear substitution algorithms that were previously unavailable to
    R users, and is most useful when other numerical,
    symbolic or algorithmic methods are impractical, inefficient or
    unavailable. }

@@

<<setup1, echo=FALSE, cache=FALSE>>=
suppressPackageStartupMessages(require(Matrix))
knitr::render_sweave()
knitr::opts_chunk$set(prompt=TRUE, cache=TRUE)
options(replace.assign=TRUE, width=90,prompt="R> ")
@

@@

Before describing how to use the package, we
present two short background notes.  The first note is an informal
mathematical explanation of numerical estimation of the Hessian
matrix, with an illustration of how the number of gradient estimates
can be reduced by exploiting the sparsity pattern and symmetric
structure.  This note borrows heavily from, and use the notation in,
\citet[Chapter 6]{MagnusNeudecker2007}. The
second note is a summary of some of the sparse matrix classes that are
defined in the \pkg{Matrix} package \citep{R_Matrix}, which are used
extensively in \pkg{sparseHessianFD}.

@@

The distinction among sparse matrix classes is important because
\pkg{sparseHessianFD}'s \method{hessian} method returns a
\class{dgCMatrix}, even though the Hessian is symmetric. Depending on how the
Hessian is used, it might be useful to coerce the Hessian into a
\class{dsCMatrix} object.  Also, the utility functions in Table~\ref{tab:helpers} expect or return certain classes of matrices, so
some degree of coercion of input and output might be
necessary. Another useful \pkg{Matrix} function is \func{tril}, which
extracts the lower triangle of a general or symmetric matrix.

@@

In both cases, the number of non-zeros is the same.   There are
\Sexpr{nels1} elements in this symmetric matrix, but only \Sexpr{nnz1} are
non-zero, and only \Sexpr{nnz1LT} values are unique.  Although the reduction in
RAM from using a sparse matrix structure for the Hessian may be
modest, consider what would happen if $N=\Sexpr{scales::comma(Q)}$ instead.  In that case,
there are \Sexpr{scales::comma(nv2)} variables in the problem, and more than $\Sexpr{scales::comma(floor(nels2/10^6))}$ million
elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
only \Sexpr{scales::comma(nnz2LT)} values.

@@

Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
so it is worthwhile to compare its performance to that of
\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in milliseconds) to
to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}, using a 2013-edition Mac Pro with 12
processing cores and 64GB of RAM.

@@

It should not be surprising that run times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The run time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.  The
plot of the ratio of Hessian-to-gradient run times is relatively flat as the size
of the dataset increases.  We can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

\FloatBarrier
\bibliography{./sparseHessianFD}

>>>>

\documentclass[jss]{jss}

@@

  \Abstract{
    Sparse Hessian matrices occur often
    in statistics, and their fast and accurate estimation can
    improve efficiency of numerical optimization and sampling
    algorithms. By exploiting the known sparsity pattern of a Hessian, methods in the
    sparseHessianFD package require many fewer function or gradient
    evaluations than would be required if the Hessian were treated as
    dense. The package implements established graph coloring and
    linear substitution algorithms that were previously unavailable to
    R users, and is most useful when other numerical,
    symbolic or algorithmic methods are impractical, inefficient or
    unavailable.}

@@

<<setup1, echo=FALSE, cache=FALSE>>=
suppressPackageStartupMessages(require(Matrix))
knitr::render_sweave()
knitr::opts_chunk$set(prompt=TRUE, cache=FALSE)
options(replace.assign=TRUE, width=90,prompt="R> ")
@

@@

Before describing how to use the package, we
present two short background notes.  The first note is an informal
mathematical explanation of numerical estimation of the Hessian
matrix, with an illustration of how the number of gradient estimates
can be reduced by exploiting the sparsity pattern and symmetric
structure.  This note borrows heavily from, and uses the notation in,
\citet[Chapter 6]{MagnusNeudecker2007}. The
second note is a summary of some of the sparse matrix classes that are
defined in the \pkg{Matrix} package \citep{R_Matrix}, which are used
extensively in \pkg{sparseHessianFD}.

@@

The distinctions among sparse matrix classes is important because
\pkg{sparseHessianFD}'s \method{hessian} method returns a
\class{dgCMatrix}, even though the Hessian is symmetric. Depending on how the
Hessian is used, it might be useful to coerce the Hessian into a
\class{dsCMatrix} object.  Also, the utility functions in Table~\ref{tab:helpers} expect or return certain classes of matrices, so
some degree of coercion of input and output might be
necessary. Another useful \pkg{Matrix} function is \func{tril}, which
extracts the lower triangle of a general or symmetric matrix.

@@

In both cases, the number of non-zeros is the same.   There are
\Sexpr{nels1} elements in this symmetric matrix, but only \Sexpr{nnz1} are
non-zero, and only \Sexpr{nnz1LT} values are unique.  Although the reduction in
RAM from using a sparse matrix structure for the Hessian may be
modest, consider what would happen if $N=\Sexpr{scales::comma(Q)}$ instead.  In that case,
there are \Sexpr{scales::comma(nv2)} variables in the problem, and more than $\Sexpr{scales::comma(floor(nels2/10^6))}$ million
elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
non-zero.  If we work with only the lower triangle of the Hessian, then we need to work with
only \Sexpr{scales::comma(nnz2LT)} values.

@@

Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
so it is worthwhile to compare its performance to that of
\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in milliseconds) to
to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.  The
plot of the ratio of Hessian-to-gradient run times is relatively flat as the size
of the dataset increases.  We can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

\FloatBarrier
\bibliography{sparseHessianFD}
